\begin{thebibliography}

  Akaike, H. (1974) ``A New Look at the Statistical Model
  Identification'', \emph{IEEE Transactions on Automatic Control},
  AC-19, pp.  716--23.

  Anderson, T. W. and Hsiao, C. (1981) ``Estimation of Dynamic Models
  with Error Components'', \emph{Journal of the American Statistical
    Association}, 76, pp. 598--606.

  Arellano, M. (2003) \emph{Panel Data Econometrics}, Oxford: Oxford
  University Press.

  Arellano, M. and Bond, S. (1991) ``Some Tests of Specification for
  Panel Data: Monte Carlo Evidence and an Application to Employment
  Equations'', \emph{The Review of Economic Studies}, 58, pp.
  277--297.

  Baiocchi, G. and Distaso, W. (2003) ``GRETL: Econometric software
  for the GNU generation'', \emph{Journal of Applied Econometrics},
  18, pp. 105--10.
  
  Baltagi, B. H. (1995) \emph{Econometric Analysis of Panel Data}, New
  York: Wiley.

  Baxter, M. and King, R. G. (1995) ``Measuring Business Cycles:
  Approximate Band-Pass Filters for Economic Time Series'', National
  Bureau of Economic Research, Working Paper No. 5022.

  Belsley, D., Kuh, E. and Welsch, R. (1980) \emph{Regression
    Diagnostics}, New York: Wiley.

  Berndt, E., Hall, B., Hall, R. and Hausman, J. (1974) ``Estimation
  and Inference in Nonlinear Structural Models'', \emph{Annals of
    Economic and Social Measurement}, 3/4, pp. 653--665.

  Blundell, R. and Bond S. (1998) ``Initial Conditions and Moment
  Restrictions in Dynamic Panel Data Models'', \emph{Journal of
    Econometrics}, 87, pp. 115--143.

  Bollerslev, T. and Ghysels, E. (1996) ``Periodic Autoregressive
  Conditional Heteroscedasticity'', \emph{Journal of Business and
    Economic Statistics}, 14, pp. 139--151.

  Box, G. E. P. and Jenkins, G. (1976) \textit{Time Series Analysis:
  Forecasting and Control}, San Franciso: Holden-Day.

  Box, G. E. P. and Muller, M. E.  (1958) ``A Note on the Generation
  of Random Normal Deviates'', \emph{Annals of Mathematical
    Statistics}, 29, pp. 610--11.

  Chesher, A. and Irish, M. (1987), ``Residual Analysis in the Grouped
  and Censored Normal Linear Model'', \emph{Journal of Econometrics},
  34, pp. 33--61.

  Cureton, E. (1967), ``The Normal Approximation to the Signed-Rank
  Sampling Distribution when Zero Differences are Present'',
  \emph{Journal of the American Statistical Association}, 62, pp.
  1068--1069.

  Davidson, R. and MacKinnon, J. G. (1993) \emph{Estimation and
    Inference in Econometrics}, New York: Oxford University Press.

  Davidson, R. and MacKinnon, J. G. (2004) \emph{Econometric Theory
    and Methods}, New York: Oxford University Press.

  Doornik, J. A. and Hansen, H.  (1994) ``An Omnibus Test for
  Univariate and Multivariate Normality'', working paper, Nuffield
  College, Oxford.

  Doornik, J. A. (1998) ``Approximations to the Asymptotic
  Distribution of Cointegration Tests'', \emph{Journal of Economic
    Surveys}, 12, pp. 573--93.  Reprinted with corrections in M.
  McAleer and L. Oxley \emph{Practical Issues in Cointegration
    Analysis}, Oxford: Blackwell, 1999.

  Edgerton, D. and Wells, C. (1994) ``Critical Values for the Cusumsq
  Statistic in Medium and Large Sized Samples'', \emph{Oxford Bulletin of
  Economics and Statistics}, 56, pp. 355--365.

  Fiorentini, G., Calzolari, G. and Panattoni, L. (1996) ``Analytic
  Derivatives and the Computation of GARCH Estimates'', \emph{Journal
    of Applied Econometrics}, 11, pp. 399--417.

  Frigo, M. and Johnson, S. G. (2005) ``The Design and Implementation of
  FFTW3,'' \emph{Proceedings of the IEEE 93}, 2, pp. 216--231 . Invited
  paper, Special Issue on Program Generation, Optimization, and Platform
  Adaptation.

  Goossens, M., Mittelbach, F., and Samarin, A. (1993) \emph{The
    \LaTeX\ Companion}, Boston: Addison-Wesley.

  Gourieroux, C., Monfort, A., Renault, E. and Trognon, A. (1987)
  ``Generalized Residuals'', \emph{Journal of Econometrics}, 34,
  pp. 5--32.  

  Greene, William H. (2000) \emph{Econometric Analysis}, 4th edition,
  Upper Saddle River, NJ: Prentice-Hall.

  Gujarati, Damodar N. (2003) \emph{Basic Econometrics}, 4th edition,
  Boston, MA: McGraw-Hill.

  Hall, Alastair D. (2005) \emph{Generalized Method of Moments},
  Oxford: Oxford University Press.

  Hamilton, James D. (1994) \emph{Time Series Analysis}, Princeton,
  NJ: Princeton University Press.

  Hannan, E. J. and Quinn, B. G. (1979) ``The Determination of the
    Order of an Autoregression'', \emph{Journal of the Royal
    Statistical Society}, B, 41, pp. 190--195.

  Hansen, L. P. (1982) ``Large Sample Properties of Generalized Method
  of Moments Estimation'', \emph{Econometrica}, 50, pp. 1029--1054.

  Hansen, L. P. and Singleton, K. J. (1982) ``Generalized
  Instrumental Variables Estimation of Nonlinear Rational Expectations
  Models'', \emph{Econometrica} 50, pp. 1269--1286. 

  Hausman, J. A. (1978) ``Specification Tests in Econometrics'',
  \emph{Econometrica}, 46, pp.\ 1251--1271.

  Heckman, J. (1979) ``Sample Selection Bias as a Specification Error'',
  \emph{Econometrica}, 47, pp\. 153--161.

  Hodrick, Robert and Prescott, Edward C. (1997) ``Postwar U.S. Business
  Cycles: An Empirical Investigation'', \emph{Journal of Money, Credit and
  Banking}, 29, pp. 1--16.

  Johansen, Søren (1995) \emph{Likelihood-Based Inference in
    Cointegrated Vector Autoregressive Models}, Oxford: Oxford
  University Press.

  Keane, Michael P. and Wolpin, Kenneth I. (1997) ``The Career
  Decisions of Young Men'', \emph{Journal of Political Economy}, 105,
  pp. 473--522.

  Kiviet, J. F. (1986) ``On the Rigour of Some Misspecification Tests
  for Modelling Dynamic Relationships'', \emph{Review of Economic
    Studies}, 53, pp. 241--261.

  Kwiatkowski, D., Phillips, P. C. B., Schmidt, P. and Shin, Y. (1992)
  ``Testing the Null of Stationarity Against the Alternative of a Unit
  Root: How Sure Are We That Economic Time Series Have a Unit Root?'',
  \emph{ Journal of Econometrics}, 54, pp. 159--178.

  Locke, C. (1976) ``A Test for the Composite Hypothesis that a
  Population has a Gamma Distribution'', \emph{Communications in
    Statistics --- Theory and Methods}, A5(4), pp. 351--364.
  
  Lucchetti, R., Papi, L., and Zazzaro, A. (2001) ``Banks'
  Inefficiency and Economic Growth: A Micro Macro Approach'',
  \emph{Scottish Journal of Political Economy}, 48, pp. 400--424.

  McCullough, B. D. and Renfro, Charles G. (1998) ``Benchmarks and
  software standards: A case study of GARCH procedures'',
  \emph{Journal of Economic and Social Measurement}, 25, pp. 59--71.

  MacKinnon, J. G. (1996) ``Numerical Distribution Functions for Unit
  Root and Cointegration Tests'', \emph{Journal of Applied
    Econometrics}, 11, pp. 601--618.
  
  MacKinnon, J. G. and White, H.  (1985) ``Some
  Heteroskedasticity-Consistent Covariance Matrix Estimators with
  Improved Finite Sample Properties'', \emph{Journal of Econometrics},
  29, pp. 305--25.
  
  Maddala, G. S. (1992) \emph{Introduction to Econometrics}, 2nd
  edition, Englewood Cliffs, NJ: Prentice-Hall.
  
  Matsumoto, M. and Nishimura, T.  (1998) ``Mersenne twister: a
  623-dimensionally equidistributed uniform pseudo-random number
  generator'', \emph{ACM Transactions on Modeling and Computer
    Simulation}, 8, pp. 3--30.

  Mroz, T. (1987) ``The Sensitivity of an Empirical Model of Married
  Women's Hours of Work to Economic and Statistical Assumptions''
  \emph{Econometrica} 55, pp. 765--799.

  Nerlove, M, (1999) ``Properties of Alternative Estimators of Dynamic
  Panel Models: An Empirical Analysis of Cross-Country Data for the
  Study of Economic Growth'', in Hsiao, C., Lahiri, K., Lee, L.-F. and
  Pesaran, M. H. (eds) \emph{Analysis of Panels and Limited Dependent
    Variable Models}, Cambridge: Cambridge University Press.
    
  Neter, J. Wasserman, W. and Kutner, M. H. (1990) \emph{Applied
    Linear Statistical Models}, 3rd edition, Boston, MA: Irwin.
  
  R Core Development Team (2000) \emph{An Introduction to R}, version
  1.1.1.

  Ramanathan, Ramu (2002) \emph{Introductory Econometrics with
    Applications}, 5th edition, Fort Worth: Harcourt.
  
  Schwarz, G. (1978) ``Estimating the dimension of a model'',
  \textit{Annals of Statistics}, 6, pp. 461--64.
  
  Shapiro, S. and Chen, L. (2001) ``Composite Tests for the Gamma
  Distribution'', \emph{Journal of Quality Technology}, 33, pp.
  47--59.
    
  Silverman, B. W. (1986) \emph{Density Estimation for Statistics and
    Data Analysis}, London: Chapman and Hall.
  
  Stock, James H. and Watson, Mark W. (2003) \emph{Introduction to
    Econometrics}, Boston, MA: Addison-Wesley.

  Swamy, P. A. V. B. and Arora, S. S. (1972) ``The Exact Finite Sample
  Properties of the Estimators of Coefficients in the Error Components
  Regression Models'', \emph{Econometrica}, 40, pp. 261--75.  

  White, H. (1980) ``A Heteroskedasticity-Consistent Covariance Matrix
  Astimator and a Direct Test for Heteroskedasticity'',
  \emph{Econometrica}, 48, pp. 817--38.

  Windmeijer, F. (2005) ``A Finite Sample Correction for the Variance
  of Linear Efficient Two-step GMM Estimators'', \emph{Journal of
    Econometrics}, 126, pp. 25--51.
  
  Wooldridge, Jeffrey M. (2002a) \emph{Econometric Analysis of Cross
    Section and Panel Data}, Cambridge, Mass.: MIT Press.

  Wooldridge, Jeffrey M. (2002b) \emph{Introductory Econometrics, A
    Modern Approach}, 2nd edition, Mason, Ohio: South-Western.

\end{thebibliography}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "gretl-guide"
%%% End: 

