\chapter{Gretl commands}
\label{cmdref}

\section{Introduction}
\label{cmd-intro}

The commands defined below may be executed interactively in the
command-line client program or in the console window of the GUI
program.  They may also be placed in a ``script'' or batch file for
non-interactive execution.
    
The following notational conventions are used below:
    
\begin{itemize}

\item A \texttt{typewriter font} is used for material that you would
  type directly, and also for internal names of variables.

\item Terms in a \textsl{slanted font} are place-holders: you should
  substitute some specific replacement.  For example, you might type
  \texttt{income} in place of the generic \textsl{xvar}.

\item The construction \texttt{[} \textsl{arg} \texttt{]} means that
  the argument \textsl{arg} is optional: you may supply it or not (but
  in any case don't type the brackets).

\item The phrase ``estimation command'' means a command that generates
  estimates for a given model, for example \cmd{ols}, \cmd{ar} or
  \cmd{wls}.

\end{itemize}

In general, each line of a command script should contain one and only
one complete gretl command.  There are, however, means of continuing a
long command from one line of input to another.  First, if the last
non-space character on a line is a backslash, this is taken to indicate
that the command is continued on the following line.  In addition, if
comma or semicolon is a valid character in context then a trailing comma
or semicolon also indicates continuation.\footnote{There's one
  exception: in the \texttt{smpl} command, a trailing semicolon
  indicates that the end of the sample range should remain unchanged; in
  that context it is not taken to call for continuation.}  To emphasize
the point: a backslash may be inserted ``arbitrarily'' to indicate
continuation, but comma and semicolon work in this capacity only if (a)
they are syntactically valid as part of the command but (b) they are not
legitimate as the terminating character of the command.

\section{Commands}
\label{cmd-cmd}

%% auto-generated from XML base, gretl_commands.xml
\input{refbody}

\section{Commands by topic}
\label{cmd-topics}

The following sections show the available commands grouped by topic.

%% auto-generated from gretl sources
\input{cmdtopics}

\section{Short-form command options}
\label{cmd-opt-short}

\input{optshort}



