\chapter{User-written functions}
\label{chap:user-funcs}

Hansl natively provides a reasonably wide array of pre-defined
functions for manipulating variables of all kinds; the previous
chapters contain several examples. However, it is also possible to
extend hansl's native capabilities by defining additional
functions.

Here's what a user-defined function looks like:
\begin{flushleft}
\texttt{function \emph{type} \emph{funcname}(\emph{parameters})}\\
   \quad \texttt{\emph{function body}}\\
\texttt{end function}
\end{flushleft}

The opening line of a function definition contains these elements, in
strict order:

\begin{enumerate}
\item The keyword \texttt{function}.
\item \texttt{\emph{type}}, which states the type of value returned by
  the function, if any.  This must be one of \texttt{void} (if the
  function does not return anything), \texttt{scalar},
  \texttt{series}, \texttt{matrix}, \texttt{list}, \texttt{string},
  \texttt{bundle}, or one of the array types, that is \texttt{bundles},
  \texttt{lists}, \texttt{matrices} and \texttt{strings};
\item \texttt{\emph{funcname}}, the unique identifier for the
  function.  Function names have a maximum length of 31 characters;
  they must start with a letter and can contain only letters, numerals
  and the underscore character. They cannot coincide with the names of
  native commands or functions.
\item The function's \texttt{\emph{parameters}}, in the form of a
  comma-separated list enclosed in parentheses. Note: parameters are
  the only way hansl function can receive anything from ``the
  outside''. In hansl there are no global variables.
\end{enumerate}

Function parameters can be of any of the types shown below.

\begin{center}
\begin{tabular}{ll}
  \multicolumn{1}{c}{Type} &
  \multicolumn{1}{c}{Description} \\ [4pt]
  \texttt{bool}   & scalar variable acting as a Boolean switch \\
  \texttt{int}    & scalar variable acting as an integer  \\
  \texttt{scalar} & scalar variable \\
  \texttt{series} & data series (see section~\ref{sec:series})\\
  \texttt{list}   & named list of series  (see section~\ref{sec:lists})\\
  \texttt{matrix} & matrix or vector \\
  \texttt{string} & string variable or string literal \\
  \texttt{bundle} & all-purpose container \\
  \texttt{matrices} & array of matrices \\
  \texttt{bundles}  & array of bundles \\
  \texttt{strings}  & array of strings \\
  \texttt{arrays}   & array of arrays \\
\end{tabular}
\end{center}

Each element in the listing of parameter must include two terms: a
type specifier, and the name by which the parameter shall be known
within the function.

The \emph{function body} contains (almost) arbitrary hansl code, which
should compute the \emph{return value}, that is the value the function
is supposed to yield. Any variable declared inside the function is
\emph{local}, so it will cease to exist when the function ends.

The \cmd{return} command is used to stop execution of the code inside
the function and deliver its result to the calling code. This
typically happens at the end of the function body, but doesn't have
to. The function definition must end with the expression
\verb|end function|, on a line of its own.

\tip{Beware: unlike some other languages (e.g.\ Matlab or GAUSS), you
  cannot directly return multiple outputs from a function. However,
  you can return a multiple-item object, such as an array for
  homogenous returns or a bundle for heterogenous items, and stuff it
  with as many objects as you want.}

In order to get a feel for how functions work in practice, here's a
simple example:
\begin{code}
function scalar quasi_log (scalar x)
   /* popular approximation to the natural logarithm
      via Padé polynomials
   */
   if x < 0
      scalar ret = NA
   else
      scalar ret = 2*(x-1)/(x+1)
   endif
   return ret
end function

loop for (x=0.5; x<2; x+=0.1)
   printf "x = %4.2f; ln(x) = %g, approx = %g\n", x, ln(x), quasi_log(x)
endloop
\end{code}

The code above computes the rational function
\[
  f(x) = 2 \cdot \frac{x-1}{x+1} ,
\]
which provides a decent approximation to the natural logarithm in the
neighborhood of 1. Some comments on the code:

\begin{enumerate}
\item Since the function is meant to return a scalar, we put the
  keyword \texttt{scalar} after the intial \cmd{function}.
\item In this case the parameter list has only one element: it is
  named \texttt{x} and is specified to be a scalar.
\item On the next line the function definition begins; the body
  includes a comment and an \cmd{if} block.
\item The function ends by returning the computed value, \texttt{ret}.
\item The lines below the function definition give a simple example of
  usage. Note that in the \cmd{printf} command, the two functions
  \cmd{ln()} and \dtk{quasi_log()} behave in exactly the same way
  from a purely syntactic viewpoint, although the former is native and
  the latter is user-defined.
\end{enumerate}

In ambitious uses of hansl you may end up writing several functions,
some of which may be quite long. In order to avoid cluttering your
script with function definitions, hansl provides the \cmd{include}
command: you can put your function definitions in a separate file (or
set of files) and read them in as needed.  For example, suppose you
saved the definition of \verb|quasi_log()| in a separate file called
\verb|quasilog_def.inp|: the code above could then be written more
compactly as
\begin{code}
include quasilog_def.inp

loop for (x=0.5; x<2; x+=0.1)
   printf "x = %4.2f; ln(x) = %g, approx = %g\n", x, ln(x), quasi_log(x)
endloop
\end{code}
Moreover, \cmd{include} commands can be nested.


\section{Parameter passing and return values}
\label{sec:params-returns}

In hansl, parameters are by default passed \emph{by value}, so what is
used inside the function is a copy of the original argument. You may
modify it, but you'll be just modifying the copy. The following
example should make this point clear:
\begin{code}
function void f(scalar x)
    x = x*2
    print x
end function

scalar x = 3
f(x)
print x
\end{code}
Running the above code yields
\begin{code}
              x =  6.0000000
              x =  3.0000000
\end{code}
The first \cmd{print} statement is executed inside the function, and
the displayed value is 6 because the input \verb|x| is doubled;
however, what really gets doubled is simply a \emph{copy} of the
original \texttt{x}: this is demonstrated by the second \cmd{print}
statement.  If you want a function to modify its arguments, you must
use pointers.

Copying the content of the incoming parameter to a local version may
have a sizeable impact on compute speed and memory usage when the
object is large (say, a 1000 $\times$ 1000 matrix). To avoid this cost
you can prepend the \texttt{const} modifier to the parameter type,
thereby promising that the object in question will not be modified
inside the function. In that case gretl will grant read-only access to
the object at caller level instead of copying it (and will flag an
error at any attempt to modify the object). Consult the \GUG{} for
further details.

\subsection{Pointers}

Each of the type-specifiers, with the exception of \texttt{list}, may
be modified by prepending an asterisk to the associated parameter
name, as in
%
\begin{code}
function scalar myfunc (matrix *y)
\end{code}
This indicates that the required argument is not a plain matrix but
rather a \emph{pointer-to-matrix}, or in other words the memory
address at which the variable is stored.

This can seem a bit mysterious to people unfamiliar with the C
programming language, so allow us to explain how pointers work by
analogy. Suppose you set up a barber shop. Ideally, your customers
would walk into your shop, sit on a chair and have their hair trimmed
or their beard shaved. However, local regulations forbid you to modify
anything coming in through your shop door. Of course, you wouldn't do
much business if people must leave your shop with their hair
untouched. Nevertheless, you have a simple way to get around this
limitation: your customers can come to your shop, tell you their home
address and walk out. Then, nobody stops you from going to their place
and exercising your fine profession. You're OK with the law, because
no modification of anything took place inside your shop.

While our imaginary restriction on the barber seems arbitrary, the
analogous restriction in a programming context is not: it prevents
functions from having unpredictable side effects. (You might be upset
if it turned out that your person was modified after visiting the
grocery store!)

In hansl (unlike C) you don't have to take any special care within the
function to distinguish the variable from its address,\footnote{In C,
  this would be called \emph{dereferencing} the pointer. The
  distinction is not required in hansl because there is no equivalent
  to operating on the supplied address itself, as in C.} you just use
the variable's name.  In order to supply the address of a variable
when you invoke the function, you use the ampersand (\verb|&|)
operator.

An example should make things clearer. The following code
\begin{code}
function void swap(scalar *a, scalar *b)
    scalar tmp = a
    a = b
    b = tmp
end function

scalar x = 0
scalar y = 1000000
swap(&x, &y)
print x y
\end{code}
gives the output
\begin{code}
              x =  1000000.0
              y =  0.0000000
\end{code}
So \texttt{x} and \texttt{y} have in fact been swapped. How?

First you have the function definition, in which the arguments are
pointers to scalars. Inside the function body, the distinction is
moot, as \verb|a| is taken to mean ``the scalar that you'll find at
the address given by the first argument'' (and likewise for
\verb|b|). The rest of the function simply swaps \texttt{a} and
\texttt{b} by means of a local temporary variable.

Outside the function, we first initialize the two scalars \texttt{x}
to 0 and \texttt{y} to a big number. When the function is called, it
is given as arguments \verb|&a| and \verb|&b|, which hansl identifies
as ``the address of'' the two scalars \texttt{a} and \texttt{b},
respectively.

Besides making it possible to modify function arguments in such a way
that they stay modified at caller level, use of pointer arguments
avoids the computational cost of copying arguments. However, it is not
idiomatic in hansl to use the pointer-argument mechanism for
cost-saving alone since the same effect can be achieved via the
\texttt{const} modifier described above.

\subsection{Advanced parameter passing and optional arguments}

The parameters to a hansl function can be also specified in more
sophisticated ways than outlined above. There are three additional
features worth mentioning:
\begin{enumerate}
\item A descriptive string can be attached to each parameter for GUI usage.
\item For some parameter types, there is a special syntax construct
  for ensuring that its value is bounded; for example, you can
  stipulate a scalar argument to be positive, or constrained within a
  pre-specified range.
\item Some of the arguments can be made optional.
\end{enumerate}

A thorough discussion is too long to fit in this document, and the
interested reader should refer to the ``User-defined functions''
chapter of the \GUG. Here we'll just show you a simple, and hopefully
self-explanatory, example which combines features 2 and 3. Suppose
you have a function for producing smileys, defined as
\begin{code}
function void smileys(int times[0::1], bool frown[0])
    if frown
        string s = ":-("
    else
        string s = ":-)"
    endif

    loop times
        printf "%s ", s
    endloop

    printf "\n"
end function
\end{code}

Then, running
\begin{code}
smileys()
smileys(2, 1)
smileys(4)
\end{code}

produces

\begin{code}
:-)
:-( :-(
:-) :-) :-) :-)
\end{code}

\subsection{Embedding arguments in bundles}

Some complex functions may require a large number of arguments. There is no
limit to the number of arguments a function can have, but an overly
complicated function signature is not pleasant to use. Some programming
languages (\app{R}, for one) obviate this problem by using \emph{named
  arguments}, so that you may call a function by supplying only the few
arguments you actually need, leaving the other ones to their default values.

In hansl we don't have named arguments, but a commonly employed technique
achieves the same result in a similar way: the idea is to package arguments
into a bundle (see Section~\ref{sec:bundles}) and use the bundle syntax to
handle its contents.

For example, say you want to write a function to extract a substring from a
string and optionally capitalize it. You could start from something like
\begin{code}
function string Sub(string s, scalar ini, scalar fin, bool capital)
    string ret = s[ini:fin]
    return capital ? toupper(ret) : ret
end function
\end{code}
so the call \texttt{Sub("nowhere", 4, 7, 1)} would produce the string
\texttt{HERE}. A more sophisticated version of the function may have default
values, so that you could call the function in a simplified form. Using the
syntax shown in the previous subsection, one could set the defaults as
\begin{code}
function string Sub(string s, scalar ini[1], scalar fin[3], bool capital[FALSE])
    string ret = s[ini:fin]
    return capital ? toupper(ret) : ret
end function
\end{code}
and the call \texttt{Sub("nowhere")} would return the string
\texttt{now}. However, if we wanted the string to be capitalized, we would
have to set the fourth parameter to 1: to get \texttt{NOW} the function
would have to be called as \texttt{Sub("nowhere", , ,1)}. With more than 5
or 6 parameters in the function signature, this becomes quite awkward.

This issue can be resolved by putting arguments 2 to 4 into a bundle, as in
\begin{code}
function string Sub(string s, bundle opts)
    string ret = s[opts.ini:opts.fin]
    return opts.capital ? toupper(ret) : ret
end function
\end{code}
where the returned string is computed using the bundle contents, and you may
call the function as
\begin{code}
  bundle myopts = _(ini=1, fin=3, capital=TRUE)
  string out = Sub("nowhere", myopts)
\end{code}
Note that these two lines could be combined as
\begin{code}
  string out = Sub("nowhere", _(ini=1, fin=3, capital=TRUE))
\end{code}
but in some cases it may be convenient to have the options bundle as a
persistent object, so that successive calls to the function may take
place with incremental changes.

This sort of mechanism lends itself naturally to handling default values in
an elegant way. Consider the code below:
\begin{code}
function string Sub(string s, bundle opts_in[null])
    bundle opts = _(ini=1, fin=3, capital=0)
    if exists(opts_in)
        opts = opts_in + opts
    endif
    string ret = s[opts.ini:opts.fin]
    return opts.capital ? toupper(ret) : ret
end function
\end{code}
Let's analyse the body of the function line by line:
\begin{enumerate}
\item the function signature contains only two arguments: the string
  to process and a bundle, which has a default value of \texttt{null}
  and so can be omitted.
\item A bundle \texttt{opts} is defined with default values for the scalars
  \texttt{ini} and \texttt{fin} and for the boolean flag \texttt{capital}.
\item If a bundle was passed as the second argument, then the line
  \begin{code}
     opts = opts_in + opts
  \end{code}
  replaces the keys in \texttt{opts} with those present in
  \dtkttt{opts_in} (with the \texttt{+} operator, the left-hand
  bundle takes precedence). At this point, the bundle \texttt{opts}
  will contain a mixture of default and user-set keys.
\item From here on, everything proceeds as above.
\end{enumerate}

This means that the call \texttt{Sub("nowhere")} would yield \texttt{now},
but if we want capitalized output we can call the function as
\begin{code}
  string out = Sub("nowhere", _(capital=TRUE))
\end{code}

The ``incremental variations'' idea for the options bundle (mentioned
above) can be now exploited as in the following code:
\begin{code}
  bundle myopts = _(capital=TRUE)
  string out1 = Sub("nowhere", myopts)
  myopts.fin = 2
  string out2 = Sub("nowhere", myopts)
\end{code}
Execution this code gives strings \texttt{out1} containing \texttt{NOW} and
\texttt{out2} containing \texttt{NO}.

At this point we have something virtually equivalent to named arguments.
Note that the keys in \verb|_()|, unlike individual function arguments, can
be given in any order.

\section{Recursion}

Hansl functions can be recursive; what follows is the obligatory
factorial example:
\begin{code}
function scalar factorial(scalar n)
    if (n<0) || (n>floor(n))
        # filter out everything that isn't a
        # non-negative integer
        return NA
    elif n==0
        return 1
    else
        return n * factorial(n-1)
    endif
end function

loop i = 0 .. 6
    printf "%d! = %d\n", i, factorial(i)
endloop
\end{code}

This is fun, but in practice you'll be much better off using the
pre-cooked gamma function (or better still, its logarithm).

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "hansl-primer"
%%% End:
