\chapter{Persistent objects}
\label{chap-persist}

%%% Work in progress, not really ready for translation yet.

FIXME This chapter needs to deal with saving models too.

\section{Named lists}
\label{named-lists}

Many \app{gretl} commands take one or more lists of variables as
arguments.  To make this easier to handle in the context of command
scripts, and in particular within user-defined functions, \app{gretl}
offers the possibility of \textit{named lists}.  

\subsection{Creating and modifying named lists}

A named list is created using the keyword \texttt{list}, followed by
the name of the list, an equals sign, and either \texttt{null} (to
create an empty list) or one or more variables to be placed on the
list.  For example,
%
\begin{code}
list xlist = 1 2 3 4
list reglist = income price 
list empty_list = null
\end{code}

The name of the list must start with a letter, and must be composed
entirely of letters, numbers or the underscore character.  The maximum
length of the name is 15 characters; list names cannot contain
spaces.  When adding variables to a list, you can refer to them either
by name or by their ID numbers. 

Once a named list has been created, it will be ``remembered'' for the
duration of the \app{gretl} session, and can be used in the context of
any \app{gretl} command where a list of variables is expected.  One
simple example is the specification of a list of regressors:
%
\begin{code}
list xlist = x1 x2 x3 x4
ols y 0 xlist
\end{code}

Lists can be modified in two ways.  To \textit{redefine} an existing
list altogether, use the same syntax as for creating a list.  For
example
%
\begin{code}
list xlist = 1 2 3
list xlist = 4 5 6
\end{code}

After the second assignment, \texttt{xlist} contains just variables 4,
5 and 6.

To \textit{append} or \textit{prepend} variables to an existing list,
we simply make use of the fact that a named list can stand in for a
``longhand'' list.  For example, we can do
%
\begin{code}
list xlist = xlist 5 6 7
list xlist = 9 10 xlist 11 12
\end{code}

\subsection{Querying a list}

You can determine whether an unknown variable actually represents a list
using the function \texttt{islist()}.
%
\begin{code}
series xl1 = log(x1)
series xl2 = log(x2)
list xlogs = xl1 xl2
genr is1 = islist(xlogs)
genr is2 = islist(xl1)
\end{code}

The first \texttt{genr} command above will assign a value of 1 to
\texttt{is1} since \texttt{xlogs} is in fact a named list.  The second
genr will assign 0 to \texttt{is2} since \texttt{xl1} is a data
series, not a list.  

You can also determine the number of variables or elements in a list
using the function \texttt{nelem()}.
%
\begin{code}
list xlist = 1 2 3
genr nl = nelem(xlist)
\end{code}

The scalar \texttt{nl} will be assigned a value of 3 since
\texttt{xlist} contains 3 members.

You can display the membership of a named list as illustrated in this
interactive session:
%
\begin{code}
? list xlist = x1 x2 x3
Added list 'xlist'
? list xlist print
 xlist: x1 x2 x3
\end{code}
%
Note that \texttt{print xlist} will do something different, namely
print the values of all the variables in \texttt{xlist} (as should be
expected).

\subsection{Generating lists of transformed variables}

Given a named list of variables, you are able to generate lists of
transformations of these variables using a special form of the
commands \texttt{logs}, \texttt{lags}, \texttt{diff}, \texttt{ldiff},
\texttt{sdiff} or \texttt{square}.  In this context these keywords
must be followed directly by a named list in parentheses.  For example
%
\begin{code}
list xlist = x1 x2 x3
list lxlist = logs(xlist)
list difflist = diff(xlist)
\end{code}

When generating a list of \textit{lags} in this way, you can specify
the maximum lag order inside the parentheses, before the list name and
separated by a comma.  For example
%
\begin{code}
list xlist = x1 x2 x3
list laglist = lags(2, xlist)
\end{code}
%
or
%
\begin{code}
scalar order = 4
list laglist = lags(order, xlist)
\end{code}

These command will populate \texttt{laglist} with the specified number
of lags of the variables in \texttt{xlist}.  (As with the ordinary
\texttt{lags} command, you can omit the order, in which case this is
determined automatically based on the frequency of the data.)  One
further special feature is available when generating lags, namely, you
can give the name of a single variable in place of a named list on the
right-hand side, as in
%
\begin{code}
series lx = log(x)
list laglist = lags(4, lx)
\end{code}

Note that the ordinary syntax for, e.g., \texttt{logs}, is just
%
\begin{code}
logs x1 x2 x3
\end{code}
%
If \texttt{xlist} is a named list, you can also say
%
\begin{code}
logs xlist
\end{code}
%
but this form will not save the logs as a named list; for that you
need the form
%
\begin{code}
list loglist = logs(xlist)
\end{code}










    
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "gretl-guide"
%%% End: 

